Demes: a standard format for demographic models

### 2022-08-01 response-to-reviewers.tex

I have found the paper clearly written and I have just a few comments,
regarding the structure of the paper and the Demes implementation itself.
\end{point}
\begin{reply}
TODO
\end{reply}

\begin{point}
I have two major concerns. The first one is about the two formats: HDM and MDM.
Contrary to what the author say, I have found the HDM more confusing to read
than the MDM, which is much more explicit (see minor comments below). My
impression is that naive users would more easily create mis-specified models in
HDM than MDM, as HDM is using a lot of implicit definitions (to avoid
redundancy). My relative extensive exposure to the feedback of users reporting
bugs for my own software tells me that users do not carefully check the
structure of their demographic models, and explicit definitions help them in
making it right. Related to that, it is unclear if the default format should
always be the HDM, which would each time need to be parsed into MDM or if
programs could use the MDM as the default input file format. Having both
possibilities would be great, but personally I would just remove the HDM
format, and add some of its nice features to the MDM (e.g. defaults)
\end{point}
\begin{reply}

We've spent considerable effort trying to anticipate ways in which users
could create erroneous demographic models. By design the Demes spec
makes many classes of model misspecification impossible to write,
it requires that errors are raised when models are ambiguous,
and it explicitly documents the required behavior for edge cases.
We honestly believe that the Demes format is easier to write
and interpret than the status quo.
However, we agree that there are some aspects of the HDM that will be
confusing for new users.

The HDM was explicitly designed to be a less-strict version of the MDM,
and our initial approach was to simply allow the omission of some fields
when there was an obvious default.
The result is that an MDM model is a valid HDM model (but not vice versa),
so users are free to be more explicit in any part of a model.
This point was not previously mentioned in the manuscript, so we have
added a paragraph to the Parsers section which explains this.

@@

We provide the following resources to assist users when writing models.
\begin{itemize}
\item
The Demes tutorial
(\url{https://popsim-consortium.github.io/demes-spec-docs/main/tutorial.html})
is a gentle introduction to writing Demes models.
It covers 99\% of anticipated uses and assumes very little about the
reader's knowledge of population genetics.
\item
The C and Python parsers have command line interfaces that take HDM as
input and can output an MDM.
So if a user is in doubt about the interpretation of some aspect of the HDM,
then they can easily check how their model will be interpreted.
\item
Demes models can be visualised using DemesDraw
(\url{https://github.com/grahamgower/demesdraw}),
which provides a sketch of how demes are related and their relative sizes.
\end{itemize}

Finally, there is no requirement for simulation software to include
an HDM to MDM parser. As an example, the demes-slim
(\url{https://github.com/grahamgower/demes-slim/})
converter only accepts MDM models (and only in JSON format).
It thus relies on users converting HDM models using an
external parser, which is a small inconvenience for a large gain
in interoperability.
This is not ideal, however, because the software must then
*validate* the model itself---validation is usually the job of the parser,
and it's undesirable for users to be manually writing or modifying models
unless they are subsequently validated according to the spec.

### 2022-08-01 paper.tex
clarify that MDM is valid HDM

>>>>

%\subsection*{Machine Data Model}
A Demes model in MDM form is also in valid HDM form, so parsing an already
resolved model does not change the model. Furthermore, this allows users to be
more verbose and write values even when they are not required by the HDM.
Being explicit may be preferred for reasons of clarity, or if the user is
unsure about how an HDM model will be resolved into an MDM model.
Although we note that in the latter case, users can simply look at the
parser output.
